---
title: "Clustering PSI"
author: "Aïsha Schuhegger"
note: "written with the help of Quirin Manz"
output: html_document
date: "2023-04-10"
---

# Binarization of the PSI value 

This file has the aim to cluster the events regarding their PSI values to be able to binarize the PSI value into 0 and 1. 
0 signifies that the exon is excluded whereby 1 signifies that the exon is included.

For that the script has different parts consisting of:
0. Libraries 
1. Read in events.csv file and create a matrix that consists events which have at least one PSI value for an IHECRE entry (no complete NA rows anymore)
2. Create histograms for all the events 
3. K means clustering is used to assign events to a cluster 
4. Visualization of k means clustering results 
5. Use a distance function (earth movers distance - emd) to calculate the distances of the histograms and create a distance matrix (emd_matrix) 
6. Hierarchical clustering for the emd matrix


# 0. Libraries
```{r}
library(data.table)
library(R.utils)
library(pbmcapply)
library(ggplot2)
library(dplyr)
library(sqldf)
library(tidyverse)
library('factoextra')
library(stats)
Rcpp::cppFunction('double emdC(NumericVector a, NumericVector b) {
  int n = a.size();
  NumericVector dist = NumericVector(n);
  double emd = 0;
  for(int i = 0; i < (n - 1); ++i) {
    dist[i + 1] = a[i] - b[i] + dist[i];
  }
  dist = abs(dist);
  for (auto& d : dist)
    emd += d;
  return emd;
}')
```


# 1. Read in events.csv file and create a matrix that consists events which have at least one PSI value for an IHECRE entry (no complete NA rows anymore)

```{r}
# Read csv file into the data frame "events"
events <- fread("/nfs/home/students/a.schuhe/template/events.csv.gz")
# Select only the columns that start with "IHECRE"
ihec_cols <- grep("^IHECRE", names(events), value = TRUE)

# Get the amount of NA values per row
amount_na_row <- apply(events[ ,..ihec_cols], 1, function(x) sum(is.na(x)))
eventsNoNARows <- events[amount_na_row < length(ihec_cols)]

# Print results 
cat('There are', as.numeric(nrow(events)-nrow(eventsNoNARows)), 'rows(events) where only NA values exist.')
cat('Without these', as.numeric(nrow(events)-nrow(eventsNoNARows)), 'NA rows, the events file contains instead of ', as.numeric(nrow(events)), ',',as.numeric(nrow(events)), 'events where each has at least one PSI value for an IHECRE entry.')

# change rownames to names of event experiment 
eventsNoNARows <- eventsNoNARows %>% column_to_rownames(var="event_id")

# create two dataframes -> one for SE and one for RI 
events_SE <- eventsNoNARows[grepl("SE", rownames(eventsNoNARows)), ]
events_RI <- eventsNoNARows[grepl("RI", rownames(eventsNoNARows)), ] 
cat('There are', as.numeric(nrow(eventsNoNARows)), ' events where each has at least one PSI value for an IHECRE entry. They consist of', as.numeric(nrow(events_SE)), 
    'Skipped Exon (SE) events and', as.numeric(nrow(events_RI)), 'Retained Intron (RI) events. So in total ', as.numeric(nrow(events_SE)+nrow(events_RI)))
```

# 2. Create histograms for all the events 
```{r}
# create psi matrices for SE and RI 
psi_SE <- as.matrix(events_SE[,ihec_cols])
psi_RI <- as.matrix(events_RI[,ihec_cols])

# create a list with normalized histogram values for the SE events and RI events 
histo_bins = 50
histo_breaks = seq(0,1,length.out=histo_bins+1)
  
histo_list_SE <- apply(psi_SE, 1, function(x) {
  x <- x[!is.na(x)]
  dens <- hist(x, breaks = histo_breaks, plot = FALSE)$density
  dens/sum(dens)
}, simplify = FALSE)

histo_list_RI <- apply(psi_RI, 1, function(x) {
  x <- x[!is.na(x)]
  dens <- hist(x, breaks = histo_breaks, plot = FALSE)$density
  dens/sum(dens)
}, simplify = FALSE)

histo_mat_SE <- do.call(rbind, histo_list_SE) #row -> event, column -> bin
histo_mat_RI <- do.call(rbind, histo_list_RI) #row -> event, column -> bin
```

# 3. K means clustering is used to assign events to a cluster 
```{r}
# For each event the normalized histogram values are represented in the histo_mat_SE or histo_mat_RI matrix. 
# Now, the k means clustering can be used to assign events to a cluster.

# how many clusters for SE/RI (https://www.r-bloggers.com/2020/05/how-to-determine-the-number-of-clusters-for-k-means-in-r/)
# map_dbl to run many models with varying value of k (centers)

tot_withinss_SE <- map_dbl(1:10,  function(k){
  model_SE <- kmeans(x = histo_mat_SE, centers = k)
  model_SE$tot.withinss
})

tot_withinss_RI <- map_dbl(1:10,  function(k){
  model_RI <- kmeans(x = histo_mat_RI, centers = k)
  model_RI$tot.withinss
})

# Generate a data frame containing both k and tot_withinss
elbow_SE <- data.frame(
  k = 1:10,
  tot_withinss = tot_withinss_SE
)
elbow_RI <- data.frame(
  k = 1:10,
  tot_withinss = tot_withinss_RI
)

# Plot the elbow plot
elbowplot_SE <- ggplot(elbow_SE, aes(x = k, y = tot_withinss)) + geom_line(color='lightblue') + geom_point() + scale_x_continuous(breaks = 1:10) + labs(title='Elbow Plot SE', x='Number of clusters k', y='Within groups sum of squares')
elbowplot_RI <- ggplot(elbow_RI, aes(x = k, y = tot_withinss)) + geom_line(color='lightpink') + geom_point() + scale_x_continuous(breaks = 1:10) + labs(title='Elbow Plot RI', x='Number of clusters k', y='Within groups sum of squares')

pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/elbowplot_SE.pdf")
elbowplot_SE #3
dev.off()

pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/elbowplot_RI.pdf")
elbowplot_RI #3 or 5
dev.off()

elbowplot_SE_RI <- ggplot() +
  geom_line(data = elbow_SE, aes(x = k, y = tot_withinss, color = "SE"), size = 1.5) +
  geom_line(data = elbow_RI, aes(x = k, y = tot_withinss, color = "RI"), size = 1.5) +
  geom_point(data = elbow_SE, aes(x = k, y = tot_withinss), size = 2) +
  geom_point(data = elbow_RI, aes(x = k, y = tot_withinss), size = 2) +
  scale_x_continuous(breaks = 1:10) +
  scale_color_manual(values = c("lightblue", "lightpink"), labels = c("SE", "RI")) +
  labs(title='Elbow Plot for Skipped Exon and Retained Intron', x='Number of clusters k', y='Within groups sum of squares')

pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/elbowplot_SE_RI.pdf")
elbowplot_SE_RI
dev.off()
```

# Kmeans clustering 

```{r}
# Kmeans clustering 
# Cluster the data into 3 clusters
set.seed(100)
kmeans_SE <- kmeans(histo_mat_SE, centers = 3)
kmeans_RI <- kmeans(histo_mat_RI, centers = 3)

#head(kmeans_SE)
# Get the cluster assignments
cluster_assignments_SE <- kmeans_SE$cluster
cluster_assignments_RI <- kmeans_RI$cluster

#cluster_assignments_SE #6196 datapoints within the cluster
#cluster_assignments_RI #557 datapoints within the cluster -> one datapoint represents one event
```

#4. Visualization of k means clustering results
```{r}
# Which cluster has which threshold for the PSI values?

# Choose 5 random datapoints (events) of each cluster and plot their histograms 
# Skipped Exon
# subset_SE_1 <- subset(cluster_assignments_SE, kmeans_SE[["cluster"]] == 1)
# subset_SE_2 <- subset(cluster_assignments_SE, kmeans_SE[["cluster"]] == 2)
# subset_SE_3 <- subset(cluster_assignments_SE, kmeans_SE[["cluster"]] == 3)

# Skipped Exon - Cluster Indices 
cluster_SE_1_indices <- which(kmeans_SE[["cluster"]] == 1)
sample_SE_1_indices <- sample(cluster_SE_1_indices, size=10)

cluster_SE_2_indices <- which(kmeans_SE[["cluster"]] == 2)
sample_SE_2_indices <- sample(cluster_SE_2_indices, size=10)

cluster_SE_3_indices <- which(kmeans_SE[["cluster"]] == 3)
sample_SE_3_indices <- sample(cluster_SE_3_indices, size=10)

sample_SE_1 <- histo_mat_SE[sample_SE_1_indices,]
sample_SE_2 <- histo_mat_SE[sample_SE_2_indices,]
sample_SE_3 <- histo_mat_SE[sample_SE_3_indices,]


pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/sample_SE_1_barplot.pdf")
sample_SE_1_histo <- apply(sample_SE_1, 1, function(x) {
  barplot(x, plot = TRUE, xlab = 'Amount of Bins', xaxt='n', main="Random event of cluster 'Included - Skipped Exon'", width=1)
})
dev.off()


pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/sample_SE_2_barplot.pdf")
sample_SE_2_histo <- apply(sample_SE_2, 1, function(x) {barplot(x, plot = TRUE, main="Random event of cluster 'Alternatively Spliced - Skipped Exon'", xlab='Amount of Bins', width=1)}) 
dev.off()

pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/sample_SE_3_barplot.pdf")
sample_SE_3_histo <- apply(sample_SE_3, 1, function(x) {barplot(x, plot = TRUE, main="Random event of cluster 'Excluded - Skipped Exon'", xlab='Amount of Bins', width=1)}) 
dev.off()

# Quantification of the cluster
cluster_SE_1_histo <- histo_mat_SE[cluster_SE_1_indices,]
nrow(cluster_SE_1_histo)
cluster_SE_2_histo <- histo_mat_SE[cluster_SE_2_indices,]
nrow(cluster_SE_2_histo)
cluster_SE_3_histo <- histo_mat_SE[cluster_SE_3_indices,]
nrow(cluster_SE_3_histo)
combined_cluster_SE <- rbind(cluster_SE_1_histo, cluster_SE_2_histo, cluster_SE_3_histo)
nrow(combined_cluster_SE)

# Check if the data is equal -> no missing values etc. 
combined_SE_sorted <- combined_cluster_SE[order(rownames(combined_cluster_SE)), ]
histo_mat_SE_sorted <- histo_mat_SE[order(rownames(histo_mat_SE)),]

all.equal(combined_SE_sorted, histo_mat_SE_sorted) #TRUE -> use of cluster_SE_x_histo is verificated 

# Calculate the median for each event 
median_cluster_SE_1 <- apply(cluster_SE_1_histo, 1, median) #calculate the median (of psi value) for each event (each row)
median_cluster_SE_2 <- apply(cluster_SE_2_histo, 1, median)
median_cluster_SE_3 <- apply(cluster_SE_3_histo, 1, median)

data_median_cluster_SE_1 <- data.frame(Median_Value = median_cluster_SE_1)
data_median_cluster_SE_2 <- data.frame(Median_Value = median_cluster_SE_2)
data_median_cluster_SE_3 <- data.frame(Median_Value = median_cluster_SE_3)

# Create the density plot for the median values of each cluster 
pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/density_median_clusterIncluded_SE.pdf")
ggplot(data_median_cluster_SE_1, aes(x = Median_Value)) +
  geom_density(fill = "skyblue3", alpha = 0.5) +
  labs(title = "Density Plot of the Median Values of Cluster 'Included - Skipped Exon'",
       x = "Median Cluster 'Included - Skipped Exon'",
       y = "Density")
dev.off()

pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/density_median_clusterAlternativelySpliced_SE.pdf")
ggplot(data_median_cluster_SE_2, aes(x = Median_Value)) +
  geom_density(fill = "slateblue3", alpha = 0.5) +
  labs(title = "Density Plot of the Median Values of Cluster 'Alternatively Spliced - Skipped Exon'",
       x = "Median Cluster 'Alternatively Spliced - Skipped Exon'",
       y = "Density")
dev.off()

pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/density_median_clusterExcluded_SE.pdf")
ggplot(data_median_cluster_SE_3, aes(x = Median_Value)) +
  geom_density(fill = "lightpink", alpha = 0.5) +
  labs(title = "Density Plot of the Median Values of Cluster 'Excluded - Skipped Exon'",
       x = "Median Cluster 'Excluded - Skipped Exon'",
       y = "Density")
dev.off()



```


```{r}
# Retained Intron 
#subset_RI_1 <- subset(cluster_assignments_RI, kmeans_RI[["cluster"]] == 1)
#subset_RI_2 <- subset(cluster_assignments_RI, kmeans_RI[["cluster"]] == 2)
#subset_RI_3 <- subset(cluster_assignments_RI, kmeans_RI[["cluster"]] == 3)
set.seed(100)
cluster_RI_1_indices <- which(kmeans_RI[["cluster"]] == 1)
sample_RI_1_indices <- sample(cluster_RI_1_indices, size=10)

cluster_RI_2_indices <- which(kmeans_RI[["cluster"]] == 2)
sample_RI_2_indices <- sample(cluster_RI_2_indices, size=10)

cluster_RI_3_indices <- which(kmeans_RI[["cluster"]] == 3)
sample_RI_3_indices <- sample(cluster_RI_3_indices, size=10)


sample_RI_1 <- histo_mat_RI[sample_RI_1_indices,]
sample_RI_2 <- histo_mat_RI[sample_RI_2_indices,]
sample_RI_3 <- histo_mat_RI[sample_RI_3_indices,]

pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/sample_RI_1_barplot.pdf")
sample_RI_1_histo <- apply(sample_RI_1, 1, function(x) {barplot(x, plot = TRUE, main="Random event of cluster 'Excluded - Retained Intron'", xlab='Amount of Bin', width=1)})
dev.off()

pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/sample_RI_2_barplot.pdf")
sample_RI_2_histo <- apply(sample_RI_2, 1, function(x) {barplot(x, plot = TRUE, main="Random event of cluster 'Included  - Retained Intron'", xlab='Amount of Bin', width=1)})
dev.off()

pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/sample_RI_3_barplot.pdf")
sample_RI_3_histo <- apply(sample_RI_3, 1, function(x) {barplot(x, plot = TRUE, main="Random event of cluster 'Alternatively Spliced - Retained Intron'", xlab='Amount of Bin', width=1)}) 
dev.off()
```

# Visualization of k means clustering results
```{r}
# Visualize kmeans cluster


#install.packages("pak", repos = sprintf("https://r-lib.github.io/p/pak/stable/%s/%s/%s", .Platform$pkgType, R.Version()$os, R.Version()$arch))
#pak::pkg_install('factoextra')

# Create a separate custom legend using ggplot2
# color skyblue3: Included
# color slateblue3: Alternatively Spliced
# color lightpink: Excluded 

custom_legend_k3 <- ggplot() +
  geom_point(aes(x = 1, y = 1, color = factor(c("Included", "Alternatively Spliced", "Excluded"))),
             shape = 16, size = 4) +
  scale_color_manual(
    name = 'Cluster',
    labels = c('Included', 'Alternatively Spliced', 'Excluded'),
    values = c('skyblue3', 'slateblue3', 'lightpink')
  ) +
  theme_void()

# Skipped Exon
# color skyblue3: Cluster 1 SE -> Included
# color slateblue3: Cluster 2 SE -> Alternatively Spliced
# color lightpink: Cluster 3 SE -> Excluded 

kmeans_plot_SE_k3 <- fviz_cluster(
  kmeans_SE, 
  geom = "point", 
  data = histo_mat_SE, 
  main = 'Cluster for the histogram values of each SE event (3 clusters)',
  palette = c('skyblue3', 'slateblue3', 'lightpink'),
  ellipse.type = 'norm') +
  theme(legend.position = 'none') 

# Combine the cluster plot and custom legend
plot_combined_SE_k3 <- cowplot::plot_grid(kmeans_plot_SE_k3, custom_legend_k3, ncol = 2, rel_widths = c(0.8, 0.2))

pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/kmeans_SE_k3_legend.pdf')
plot_combined_SE_k3
dev.off()
```


```{r}
# Retained Intron
# color lightpink: Cluster 1 RI -> Excluded
# color skyblue3: Cluster 2 RI -> Included
# color slateblue3: Cluster 3 RI -> Alternatively Spliced

kmeans_plot_RI_k3 <- fviz_cluster(
  kmeans_RI, 
  geom = "point", 
  data = histo_mat_RI, 
  main = 'Cluster for the histogram values of each RI event (3 clusters)',
  palette = c('lightpink', 'skyblue3', 'slateblue3'),
  ellipse.type = 'norm') +
  theme(legend.position = 'none') 

# Combine the cluster plot and custom legend
plot_combined_RI_k3 <- cowplot::plot_grid(kmeans_plot_RI_k3, custom_legend_k3, ncol = 2, rel_widths = c(0.8, 0.2))

pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/kmeans_RI_k3_legend.pdf')
plot_combined_RI_k3
dev.off()


```

```{r}
# Further visualization and research if subcluster (k=4) help in splitting the 'Alternatively Spliced' Cluster in different subcluster

set.seed(100)
kmeans_SE_k4 <- kmeans(histo_mat_SE, centers = 4)
kmeans_RI_k4 <- kmeans(histo_mat_RI, centers = 4)

kmeans_plot_SE_k4 <- fviz_cluster(
  kmeans_SE_k4, 
  geom = "point", 
  data = histo_mat_SE, 
  main='Cluster for the histogram values of each SE event', 
  palette='Set2', 
  ellipse.type='norm') 

pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/kmeans_SE_k4.pdf')
kmeans_plot_SE_k4
dev.off()

kmeans_plot_RI_k4 <- fviz_cluster(
  kmeans_RI_k4, 
  geom = "point", 
  data = histo_mat_RI, 
  main='Cluster for the histogram values of each RI event', 
  palette='Set2', 
  ellipse.type='norm')

pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/kmeans_RI_k4.pdf')
kmeans_plot_RI_k4
dev.off()

```

# 5. Use a distance function (earth movers distance - emd) to calculate the distances of the histograms and create a distance matrix (emd_matrix) 

```{r}
# function: pbmclapply -> it is a wrapper around the mclapply function & it adds a progress bar to mclapply function
upper_triangle_mat_SE <- rbindlist(pbmcapply::pbmclapply(
  combn(histo_list_SE, 2, simplify=FALSE),
  function(k) {
    list(combination=paste(names(k), collapse='-'), emd=emdC(k[[1]], k[[2]]))
  }, mc.cores=20
))

upper_triangle_mat_SE[, c('from', 'to') := tstrsplit(combination, "-", fixed=TRUE, type.convert = TRUE)]
upper_triangle_mat_SE[, combination:=NULL]
lower_triangle_mat_SE <- upper_triangle[, .(emd, from=to, to=from)]
diagonal_mat_SE <- data.table(emd=0, from=events[, ID], to=events[, ID])

emd_mat_SE <- rbindlist(list(upper_triangle_mat_SE, diagonal_mat_SE, lower_triangle_mat_SE))
dist_dt_SE <- dcast(emd_mat_SE, from ~ to, value.var = "emd")
dist_SE <- as.dist(as.matrix(dist_dt_SE[, -"from"]))
plot(hclust(dist_SE))

#RI

upper_triangle_mat_RI <- rbindlist(pbmcapply::pbmclapply(
  combn(histo_list_RI, 2, simplify=FALSE),
  function(k) {
    list(combination=paste(names(k), collapse='-'), emd=emdC(k[[1]], k[[2]]))
  }, mc.cores=20
))

upper_triangle_mat_RI[, c('from', 'to') := tstrsplit(combination, "-", fixed=TRUE, type.convert = TRUE)]
upper_triangle_mat_RI[, combination:=NULL]
lower_triangle_mat_RI <- upper_triangle[, .(emd, from=to, to=from)]
diagonal_mat_RI <- data.table(emd=0, from=events[, ID], to=events[, ID])

emd_mat_rI <- rbindlist(list(upper_triangle_mat_RI, diagonal_mat_RI, lower_triangle_mat_RI))
dist_dt_RI <- dcast(emd_mat_RI, from ~ to, value.var = "emd")
dist_RI <- as.dist(as.matrix(dist_dt_RI[, -"from"]))
plot(hclust(dist_RI))
```

# 6. Hierarchical clustering for the emd matrix
```{r}
TODO
```

