---
title: "discretized_form"
output: html_document
date: "2023-05-25"
author: "Aïsha Schuhegger" 
Note: "Download and run all Rmd files from the github repository. You will need the ermerging results." 
---

```{r}
library(dplyr)
```


Create a new file (baesd on the data from the events file) where all the events belonging to the 'Included' or 'Excluded' cluster (regarding hclust) are excluded. 

```{r}
#Retained Intron 
cluster_RI_3_histo_hclust #AS cluster regarding RI (based on specific events from the events.csv file)

psi_RI_AS_1_df <- as.data.frame(psi_RI_AS_1_hclust)
psi_RI_AS_2_df <- as.data.frame(psi_RI_AS_2_hclust)
psi_RI_AS_3_df <- as.data.frame(psi_RI_AS_3_hclust)
psi_RI_AS_4_df <- as.data.frame(psi_RI_AS_4_hclust)
psi_RI_AS_5_df <- as.data.frame(psi_RI_AS_5_hclust)
psi_RI_AS_6_df <- as.data.frame(psi_RI_AS_6_hclust)

median_RI_AS_1 <- as.vector(quantile_RI_AS_1_hclust)
median_RI_AS_2 <- as.vector(quantile_RI_AS_2_hclust)
median_RI_AS_3 <- as.vector(quantile_RI_AS_3_hclust)
median_RI_AS_4 <- as.vector(quantile_RI_AS_4_hclust)
median_RI_AS_5 <- as.vector(quantile_RI_AS_5_hclust)
median_RI_AS_6 <- as.vector(quantile_RI_AS_6_hclust)

psi_RI_AS_1_df <- ifelse(psi_RI_AS_1_df > median_RI_AS_1[3], 1, 0)
psi_RI_AS_2_df <- ifelse(psi_RI_AS_2_df > median_RI_AS_2[3], 1, 0)
psi_RI_AS_3_df <- ifelse(psi_RI_AS_3_df > median_RI_AS_3[3], 1, 0)
psi_RI_AS_4_df <- ifelse(psi_RI_AS_4_df > median_RI_AS_4[3], 1, 0)
psi_RI_AS_5_df <- ifelse(psi_RI_AS_5_df > median_RI_AS_5[3], 1, 0)
psi_RI_AS_6_df <- ifelse(psi_RI_AS_6_df > median_RI_AS_6[3], 1, 0)
```

```{r}
#Retained Intron 
#cluster_RI_3_histo_hclust #AS cluster regarding RI (based on specific events from the events.csv file)

#df_names <- paste0("psi_RI_AS_", 1:6, "_df")
#hclust_names <- paste0("psi_RI_AS_", 1:6, "_hclust")
#median_names <- paste0("median_RI_AS_", 1:6)

#lapply(seq_along(df_names), function(i) {
#  psi_RI_AS_df <- get(df_names[i])
#  median_RI_AS <- get(median_names[i])
#  assign(df_names[i], as.data.frame(get(hclust_names[i])))
#  assign(median_names[i], as.vector(get(hclust_names[i])))
#  psi_RI_AS_df <- ifelse(psi_RI_AS_df > median_RI_AS[3], 1, 0)
#  assign(df_names[i], psi_RI_AS_df)
#})
```


```{r}
discretized_df_RI <- dplyr::bind_rows(as.data.frame(psi_RI_AS_1_df), as.data.frame(psi_RI_AS_2_df), as.data.frame(psi_RI_AS_3_df), as.data.frame(psi_RI_AS_4_df), as.data.frame(psi_RI_AS_5_df), as.data.frame(psi_RI_AS_6_df))
discretized_df_RI

#cluster_SE_1_histo_hclust #AS cluster regarding SE 
```

```{r}
#Skipped Exon 
cluster_SE_1_histo_hclust #AS cluster regarding SE 

psi_SE_AS_1_df <- as.data.frame(psi_SE_AS_1_hclust)
psi_SE_AS_2_df <- as.data.frame(psi_SE_AS_2_hclust)
psi_SE_AS_3_df <- as.data.frame(psi_SE_AS_3_hclust)
psi_SE_AS_4_df <- as.data.frame(psi_SE_AS_4_hclust)
psi_SE_AS_5_df <- as.data.frame(psi_SE_AS_5_hclust)
psi_SE_AS_6_df <- as.data.frame(psi_SE_AS_6_hclust)

median_SE_AS_1 <- as.vector(quantile_SE_AS_1_hclust)
median_SE_AS_2 <- as.vector(quantile_SE_AS_2_hclust)
median_SE_AS_3 <- as.vector(quantile_SE_AS_3_hclust)
median_SE_AS_4 <- as.vector(quantile_SE_AS_4_hclust)
median_SE_AS_5 <- as.vector(quantile_SE_AS_5_hclust)
median_SE_AS_6 <- as.vector(quantile_SE_AS_6_hclust)

psi_SE_AS_1_df <- ifelse(psi_SE_AS_1_df > median_SE_AS_1[3], 1, 0)
psi_SE_AS_2_df <- ifelse(psi_SE_AS_2_df > median_SE_AS_2[3], 1, 0)
psi_SE_AS_3_df <- ifelse(psi_SE_AS_3_df > median_SE_AS_3[3], 1, 0)
psi_SE_AS_4_df <- ifelse(psi_SE_AS_4_df > median_SE_AS_4[3], 1, 0)
psi_SE_AS_5_df <- ifelse(psi_SE_AS_5_df > median_SE_AS_5[3], 1, 0)
psi_SE_AS_6_df <- ifelse(psi_SE_AS_6_df > median_SE_AS_6[3], 1, 0)


discretized_df_SE <- dplyr::bind_rows(as.data.frame(psi_SE_AS_1_df), as.data.frame(psi_SE_AS_2_df), as.data.frame(psi_SE_AS_3_df), as.data.frame(psi_SE_AS_4_df), as.data.frame(psi_SE_AS_5_df), as.data.frame(psi_SE_AS_6_df))
discretized_df_SE

#cluster_SE_1_histo_hclust #AS cluster regarding SE 
```

Combine discretized_df_RI and discretized_df_SE
```{r}
discretized_df <- dplyr::bind_rows(discretized_df_RI, discretized_df_SE)
# Order the data frame based on row names
rownames_discretized_df <- as.vector(rownames(discretized_df))

# Order the data frame based on row names
discretized_df <- discretized_df[order(rownames_discretized_df), ]

#for (i in 1:nrow(discretized_df)){
#  rownames(discretized_df)[i] <- strsplit(rownames(discretized_df), ";")[[i]][1]
#}
# -> take event_id as identification for the rownames 

head(discretized_df)

```


Create a file for every row and exclude the columns that have NA as value

```{r}
library(dplyr)

epiatlas <- read.csv("/nfs/data/IHEC/RNAseq/epiatlas_metadata.csv")
helper_eventID$ID <- as.integer(helper_eventID$ID)
histone_modifications = c("H3K27ac", "H3K27me3", "H3K36me3", "H3K4me1", "H3K4me3", "H3K9me3")
for (histone_modification in histone_modifications) {
  # Create a new dataframe specific to the histone modification
  temp_df <- discretized_df
  
  i <- 1
  for (col_name in colnames(discretized_df)) {
    filtered_data <- epiatlas %>% filter(grepl(col_name, epirr_id) & experiment_type == histone_modification & assay_type == "ChIP-Seq")
    file_path <- filtered_data$data_file_path
    modified_file_path <- basename(gsub("\\*", "fc.signal", file_path))
    colnames(temp_df)[i] <- modified_file_path
    i <- i + 1
  }
  
  # Loop through each row in the new dataframe for the specific histone modification
  for (i in 1:nrow(temp_df)) {
    selected_row <- temp_df[i, ]
    selected_row <- as.data.frame(selected_row)
    #row_name <- sub(";", "_", rownames(temp_df)[i])
    #row_name <- helper_eventID$ID[helper_eventID$event_id == rownames(temp_df[i,])]
    row_name <- sprintf("%015d", helper_eventID$ID[helper_eventID$event_id == rownames(temp_df[i,])])
    rownames(selected_row) <- row_name
    
    selected_row <- selected_row[, !colSums(is.na(selected_row))]
    
    # Define the file name
    file_name <- paste0("output_discretized/", row_name, "_", histone_modification, ".txt")
    
    # Write the selected row to a file
    write.table(selected_row, file = file_name, row.names = TRUE, col.names = TRUE, sep = "\t", quote=FALSE)
  }
}

```


Now: 0.5 is the threshold for every event 
-> comparison between discretized_df and discretized_05_df
```{r}
#cluster_RI_3_histo_hclust #AS cluster regarding RI
#cluster_SE_1_histo_hclust #AS cluster regarding SE

cluster_RI_3_histo_hclust <- as.data.frame(cluster_RI_3_histo_hclust)
cluster_SE_1_histo_hclust <- as.data.frame(cluster_SE_1_histo_hclust)

psi_RI_AS_05_df <- ifelse(cluster_RI_3_histo_hclust > 0.5, 1, 0)
psi_SE_AS_05_df <- ifelse(cluster_SE_1_histo_hclust > 0.5, 1, 0)


discretized_05_df <- dplyr::bind_rows(as.data.frame(psi_RI_AS_05_df), as.data.frame(psi_SE_AS_05_df))

rownames_discretized_05_df <- as.vector(rownames(discretized_05_df))

# Order the data frame based on row names
discretized_05_df <- discretized_05_df[order(rownames_discretized_05_df), ]

```

Comparison between the data frames 

```{r}
rowSums(discretized_df == discretized_05_df, na.rm=TRUE) #-> & NA beachten 
```

