---
title: "hclust-RI"
output: html_document
date: "2023-05-11"
author: "Aïsha Schuhegger"
note: "written with the help of Quirin Manz" 
---
0. Load libraries
1. Use a distance function (earth movers distance - emd) to calculate the distances of the histograms and create a distance matrix (emd_matrix) for the event type Retained Intron
2. Use hierarchical clustering for the emd matrix


0. Load libraries
```{r}
library(data.table)
library(R.utils)
library(pbmcapply)
library(ggplot2)
library(dplyr)
library(sqldf)
library(tidyverse)
library('factoextra')
library(stats)
library(knitr)
library(kableExtra)
Rcpp::cppFunction('double emdC(NumericVector a, NumericVector b) {
  int n = a.size();
  NumericVector dist = NumericVector(n);
  double emd = 0;
  for(int i = 0; i < (n - 1); ++i) {
    dist[i + 1] = a[i] - b[i] + dist[i];
  }
  dist = abs(dist);
  for (auto& d : dist)
    emd += d;
  return emd;
}')
```

1. Use a distance function (earth movers distance - emd) to calculate the distances of the histograms and create a distance matrix (emd_matrix) for the event type Retained Intron
```{r} 
events <- fread("/nfs/home/students/a.schuhe/template/events.csv.gz")
ihecs <- names(events)[startsWith(names(events), prefix = "IHECRE")]
na_per_row <- apply(events[, ..ihecs], 1, function(x) sum(is.na(x)))
events <- events[na_per_row < length(ihecs)]
events_RI_hclust <- events[grep("RI", events$event_id), ] #only look at the Retained Intron (RI) events
#dim(events_RI_hclust)
n_bins <- 50
my_breaks <- seq(0,1,length.out=n_bins+1)
psi_mat_RI_hclust <- as.matrix(events_RI_hclust[, ..ihecs])
dens_list_RI_hclust <- apply(psi_mat_RI_hclust, 1, function(x) {
  x <- x[!is.na(x)]
  dens <- hist(x, breaks = my_breaks, plot = FALSE)$density
  dens/sum(dens)
}, simplify = FALSE)
names(dens_list_RI_hclust) <- as.character(events_RI_hclust[, as.character(ID)])
```


```{r}
#Calculate the emd matrix for the RI events 
upper_triangle_RI <- rbindlist(pbmcapply::pbmclapply(
  combn(dens_list_RI_hclust, 2, simplify = FALSE), 
  function(l) {
    list(combination = paste(names(l), collapse = "-"), emd=emdC(l[[1]], l[[2]]))
  }, mc.cores=20))
upper_triangle_RI[, c('from', 'to') := tstrsplit(combination, "-", fixed=TRUE, type.convert = TRUE)]
upper_triangle_RI[, combination:=NULL]
lower_triangle_RI <- upper_triangle_RI[, .(emd, from=to, to=from)]
diagonal_RI <- data.table(emd=0, from=events_RI_hclust[, ID], to=events_RI_hclust[, ID])
emd_mat_RI <- rbindlist(list(upper_triangle_RI, diagonal_RI, lower_triangle_RI))
dist_dt_RI <-  data.table::dcast(emd_mat_RI, from ~ to, value.var = "emd")
dist_dt_RI <- dist_dt_RI[, -"from"]
dim(dist_dt_RI)
dist_RI <- as.dist(as.matrix(dist_dt_RI))
hclust_result_RI <- hclust(dist_RI)
```

2. Use hierarchical clustering for the emd matrix

```{r}
# Plot the dendrogram with three branches
pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/dendrogram_RI.pdf")
plot(hclust_result_RI, hang = -1, main = "Dendrogram Retained Intron", labels = FALSE, xlab = "Distance")
rect.hclust(hclust_result_RI, k = 3, border = "lightpink")
dev.off()

#Look at the intersection between hclust cluster and kmeans cluster 

assigned_clusters_RI <- cutree(hclust_result_RI, k = 3)
clusters_df_RI <- data.frame(events_RI_hclust, cluster = assigned_clusters_RI)
rownames(clusters_df_RI) <- clusters_df_RI$event_id

 
cluster_RI_1_indices_hclust <- which(clusters_df_RI[["cluster"]] == 1)
cluster_RI_2_indices_hclust <- which(clusters_df_RI[["cluster"]] == 2)
cluster_RI_3_indices_hclust <- which(clusters_df_RI[["cluster"]] == 3)

events_RI_hclust <- as.data.frame(events_RI_hclust)
events_RI_hclust <- events_RI_hclust %>% column_to_rownames(var="event_id")

cluster_RI_1_histo_hclust <- events_RI_hclust[cluster_RI_1_indices_hclust,ihecs]
cluster_RI_2_histo_hclust <- events_RI_hclust[cluster_RI_2_indices_hclust,ihecs]
cluster_RI_3_histo_hclust <- events_RI_hclust[cluster_RI_3_indices_hclust,ihecs]

#rownames(cluster_RI_1_histo_hclust) <- as.character(events_RI_hclust[as.character(event_id),])
# Calculate psi matrix for each cluster 

psi_mat_RI_hclust <- as.matrix(events_RI_hclust[, ihecs])

psi_RI_1_hclust <- psi_mat_RI_hclust[rownames(clusters_df_RI[cluster_RI_1_indices_hclust, ]),]
psi_RI_2_hclust <- psi_mat_RI_hclust[rownames(clusters_df_RI[cluster_RI_2_indices_hclust, ]),]
psi_RI_3_hclust <- psi_mat_RI_hclust[rownames(clusters_df_RI[cluster_RI_3_indices_hclust, ]),]

#check number -> should be equal to dim of events_RI_hclust (for Skipped Exon)
nrow(psi_RI_1_hclust) + nrow(psi_RI_2_hclust) + nrow(psi_RI_3_hclust)
all <- rbind(psi_RI_1_hclust, psi_RI_2_hclust, psi_RI_3_hclust)
# Count the number of non-missing rows in the merged data frame

#rownames_all <- rownames(all)
#rownames_psi_RI_Included <- rownames(psi_RI_Included)
#common_row_names <- intersect(rownames_all, rownames_psi_RI_Included)
#length(common_row_names)

rownames_psi_RI_Included <- rownames(psi_RI_Included)
rownames_psi_RI_Excluded <- rownames(psi_RI_Excluded)
rownames_psi_RI_AltSpl <- rownames(psi_RI_AltSpl)

psi_RI_1_hclust <- as.data.frame(psi_RI_1_hclust)
rownames_RI_1_hclust <- rownames(psi_RI_1_hclust)

common_row_names <- intersect(rownames_RI_1_hclust, rownames_psi_RI_Included)
length(common_row_names)
common_row_names <- intersect(rownames_RI_1_hclust, rownames_psi_RI_Excluded)
length(common_row_names)
common_row_names <- intersect(rownames_RI_1_hclust, rownames_psi_RI_AltSpl)
length(common_row_names)

psi_RI_2_hclust <- as.data.frame(psi_RI_2_hclust)
rownames_RI_2_hclust <- rownames(psi_RI_2_hclust)

common_row_names <- intersect(rownames_RI_2_hclust, rownames_psi_RI_Included)
length(common_row_names)
common_row_names <- intersect(rownames_RI_2_hclust, rownames_psi_RI_Excluded)
length(common_row_names)
common_row_names <- intersect(rownames_RI_2_hclust, rownames_psi_RI_AltSpl)
length(common_row_names)


psi_RI_3_hclust <- as.data.frame(psi_RI_3_hclust)
rownames_RI_3_hclust <- rownames(psi_RI_3_hclust)

common_row_names <- intersect(rownames_RI_3_hclust, rownames_psi_RI_Included)
length(common_row_names)
common_row_names <- intersect(rownames_RI_3_hclust, rownames_psi_RI_Excluded)
length(common_row_names)
common_row_names <- intersect(rownames_RI_3_hclust, rownames_psi_RI_AltSpl)
length(common_row_names)

# Calculate the median for each event 
median_cluster_RI_1_hclust <- apply(psi_RI_1_hclust, 1, median, na.rm=TRUE) #calculate the median (of psi value) for each event (each row)
median_cluster_RI_2_hclust <- apply(psi_RI_2_hclust, 1, median, na.rm=TRUE)
median_cluster_RI_3_hclust <- apply(psi_RI_3_hclust, 1, median, na.rm=TRUE)

data_median_cluster_RI_1_hclust <- data.frame(Median_Value = median_cluster_RI_1_hclust)
data_median_cluster_RI_2_hclust <- data.frame(Median_Value = median_cluster_RI_2_hclust)
data_median_cluster_RI_3_hclust <- data.frame(Median_Value = median_cluster_RI_3_hclust)

data_median_cluster_RI_1_hclust <-  data_median_cluster_RI_1_hclust[order(data_median_cluster_RI_1_hclust$Median_Value),]
quantile_RI_1_hclust <- quantile(data_median_cluster_RI_1_hclust, probs = seq(0, 1, 1/4))
quantile_RI_1_hclust

data_median_cluster_RI_2_hclust <-  data_median_cluster_RI_2_hclust[order(data_median_cluster_RI_2_hclust$Median_Value),]
quantile_RI_2_hclust <- quantile(data_median_cluster_RI_2_hclust, probs = seq(0, 1, 1/4))
quantile_RI_2_hclust

data_median_cluster_RI_3_hclust <-  data_median_cluster_RI_3_hclust[order(data_median_cluster_RI_3_hclust$Median_Value),]
quantile_RI_3_hclust <- quantile(data_median_cluster_RI_3_hclust, probs = seq(0, 1, 1/4))
quantile_RI_3_hclust 

quantiles_RI_hclust <- data.frame(quantile_RI_2_hclust, quantile_RI_3_hclust, quantile_RI_1_hclust)
#cluster 1 -> Included
#cluster 2 -> Excluded
#cluster 3 -> Alternatively Spliced 
col_names_RI_hclust <- c("Excluded", "Alternatively Spliced", "Included")
colnames(quantiles_RI_hclust) <- col_names_RI_hclust
quantiles_RI_hclust
table_quantiles_RI_hclust <- kable(quantiles_RI_hclust, format = "html", table.attr = "border=\"1\"", 
                            caption = "Quantiles RI - hierarchical clustering", align="ccc") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
table_quantiles_RI_hclust
save_kable(x=table_quantiles_RI_hclust, file= "/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/quantiles_RI_hclust.html")

```

Having a further look on cluster 3 -> Alternatively Spliced 
```{r}
# Extract the third branch because cluster 3 -> Alternatively Spliced 
# Already done above:
subcluster_RI_3_hclust <- which(cutree_result_RI == 3)
subcluster_RI_3_hclust <- as.data.frame(subcluster_RI_3_hclust)
#nrow(subcluster_RI_3_hclust)
colnames_dist_dt_RI_3 <- c(rownames(subcluster_RI_3_hclust))
dist_dt_RI_3 <- dist_dt_RI[subcluster_RI_3_hclust$subcluster_RI_3_hclust,..colnames_dist_dt_RI_3]
#dim(dist_dt_RI_3)

dist_RI_3 <- as.dist(as.matrix(dist_dt_RI_3))
hclust_result_RI_3 <- hclust(dist_RI_3)

plot(hclust_result_RI_3, label=FALSE)

pdf("/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/dendrogram_RI_AltSpl.pdf")
plot(hclust_result_RI_3, hang = -1, main = "Dendrogram Retained Intron - only Alternatively Spliced Branch", labels = FALSE, xlab = "Distance")
rect.hclust(hclust_result_RI_3, k = 6, border = "lightpink")
dev.off()

```
```{r}
assigned_clusters_RI_AltSpl <- cutree(hclust_result_RI_3, k = 6)
events_RI_hclust_AltSpl <- events_RI_hclust[cluster_RI_3_indices_hclust,]
#nrow(events_RI_hclust_AltSpl)
clusters_df_RI_AltSpl <- data.frame(events_RI_hclust_AltSpl, cluster = assigned_clusters_RI_AltSpl)
rownames(clusters_df_RI_AltSpl) <- rownames(events_RI_hclust_AltSpl)

 
cluster_RI_AS_1_indices_hclust <- which(clusters_df_RI_AltSpl[["cluster"]] == 1)
cluster_RI_AS_2_indices_hclust <- which(clusters_df_RI_AltSpl[["cluster"]] == 2)
cluster_RI_AS_3_indices_hclust <- which(clusters_df_RI_AltSpl[["cluster"]] == 3)
cluster_RI_AS_4_indices_hclust <- which(clusters_df_RI_AltSpl[["cluster"]] == 4)
cluster_RI_AS_5_indices_hclust <- which(clusters_df_RI_AltSpl[["cluster"]] == 5)
cluster_RI_AS_6_indices_hclust <- which(clusters_df_RI_AltSpl[["cluster"]] == 6)

cluster_RI_AS_1_histo_hclust <- events_RI_hclust_AltSpl[cluster_RI_AS_1_indices_hclust,ihecs]
cluster_RI_AS_2_histo_hclust <- events_RI_hclust_AltSpl[cluster_RI_AS_2_indices_hclust,ihecs]
cluster_RI_AS_3_histo_hclust <- events_RI_hclust_AltSpl[cluster_RI_AS_3_indices_hclust,ihecs]
cluster_RI_AS_4_histo_hclust <- events_RI_hclust_AltSpl[cluster_RI_AS_4_indices_hclust,ihecs]
cluster_RI_AS_5_histo_hclust <- events_RI_hclust_AltSpl[cluster_RI_AS_5_indices_hclust,ihecs]
cluster_RI_AS_6_histo_hclust <- events_RI_hclust_AltSpl[cluster_RI_AS_6_indices_hclust,ihecs]

# Calculate psi matrix for each cluster 

psi_mat_RI_AS_hclust <- as.matrix(events_RI_hclust_AltSpl[, ihecs])

psi_RI_AS_1_hclust <- psi_mat_RI_hclust[rownames(clusters_df_RI_AltSpl[cluster_RI_AS_1_indices_hclust, ]),]
psi_RI_AS_2_hclust <- psi_mat_RI_hclust[rownames(clusters_df_RI_AltSpl[cluster_RI_AS_2_indices_hclust, ]),]
psi_RI_AS_3_hclust <- psi_mat_RI_hclust[rownames(clusters_df_RI_AltSpl[cluster_RI_AS_3_indices_hclust, ]),]
psi_RI_AS_4_hclust <- psi_mat_RI_hclust[rownames(clusters_df_RI_AltSpl[cluster_RI_AS_4_indices_hclust, ]),]
psi_RI_AS_5_hclust <- psi_mat_RI_hclust[rownames(clusters_df_RI_AltSpl[cluster_RI_AS_5_indices_hclust, ]),]
psi_RI_AS_6_hclust <- psi_mat_RI_hclust[rownames(clusters_df_RI_AltSpl[cluster_RI_AS_6_indices_hclust, ]),]

#check number -> should be equal to dim of events_RI_hclust_AltSpl (for Skipped Exon)
nrow(psi_RI_AS_1_hclust) + nrow(psi_RI_AS_2_hclust) + nrow(psi_RI_AS_3_hclust) + nrow(psi_RI_AS_4_hclust) + nrow(psi_RI_AS_5_hclust) + nrow(psi_RI_AS_6_hclust)
dim(events_RI_hclust_AltSpl)

# Calculate the median for each event 
median_cluster_RI_AS_1_hclust <- apply(psi_RI_AS_1_hclust, 1, median, na.rm=TRUE) #calculate the median (of psi value) for each event (each row)
median_cluster_RI_AS_2_hclust <- apply(psi_RI_AS_2_hclust, 1, median, na.rm=TRUE)
median_cluster_RI_AS_3_hclust <- apply(psi_RI_AS_3_hclust, 1, median, na.rm=TRUE)
median_cluster_RI_AS_4_hclust <- apply(psi_RI_AS_4_hclust, 1, median, na.rm=TRUE)
median_cluster_RI_AS_5_hclust <- apply(psi_RI_AS_5_hclust, 1, median, na.rm=TRUE)
median_cluster_RI_AS_6_hclust <- apply(psi_RI_AS_6_hclust, 1, median, na.rm=TRUE)


data_median_cluster_RI_AS_1_hclust <- data.frame(Median_Value = median_cluster_RI_AS_1_hclust)
data_median_cluster_RI_AS_2_hclust <- data.frame(Median_Value = median_cluster_RI_AS_2_hclust)
data_median_cluster_RI_AS_3_hclust <- data.frame(Median_Value = median_cluster_RI_AS_3_hclust)
data_median_cluster_RI_AS_4_hclust <- data.frame(Median_Value = median_cluster_RI_AS_4_hclust)
data_median_cluster_RI_AS_5_hclust <- data.frame(Median_Value = median_cluster_RI_AS_5_hclust)
data_median_cluster_RI_AS_6_hclust <- data.frame(Median_Value = median_cluster_RI_AS_6_hclust)

data_median_cluster_RI_AS_1_hclust <-  data_median_cluster_RI_AS_1_hclust[order(data_median_cluster_RI_AS_1_hclust$Median_Value),]
quantile_RI_AS_1_hclust <- quantile(data_median_cluster_RI_AS_1_hclust, probs = seq(0, 1, 1/4))
quantile_RI_AS_1_hclust

data_median_cluster_RI_AS_2_hclust <-  data_median_cluster_RI_AS_2_hclust[order(data_median_cluster_RI_AS_2_hclust$Median_Value),]
quantile_RI_AS_2_hclust <- quantile(data_median_cluster_RI_AS_2_hclust, probs = seq(0, 1, 1/4))
quantile_RI_AS_2_hclust

data_median_cluster_RI_AS_3_hclust <-  data_median_cluster_RI_AS_3_hclust[order(data_median_cluster_RI_AS_3_hclust$Median_Value),]
quantile_RI_AS_3_hclust <- quantile(data_median_cluster_RI_AS_3_hclust, probs = seq(0, 1, 1/4))
quantile_RI_AS_3_hclust

data_median_cluster_RI_AS_4_hclust <-  data_median_cluster_RI_AS_4_hclust[order(data_median_cluster_RI_AS_4_hclust$Median_Value),]
quantile_RI_AS_4_hclust <- quantile(data_median_cluster_RI_AS_4_hclust, probs = seq(0, 1, 1/4))
quantile_RI_AS_4_hclust

data_median_cluster_RI_AS_5_hclust <-  data_median_cluster_RI_AS_5_hclust[order(data_median_cluster_RI_AS_5_hclust$Median_Value),]
quantile_RI_AS_5_hclust <- quantile(data_median_cluster_RI_AS_5_hclust, probs = seq(0, 1, 1/4))
quantile_RI_AS_5_hclust

data_median_cluster_RI_AS_6_hclust <-  data_median_cluster_RI_AS_6_hclust[order(data_median_cluster_RI_AS_6_hclust$Median_Value),]
quantile_RI_AS_6_hclust <- quantile(data_median_cluster_RI_AS_6_hclust, probs = seq(0, 1, 1/4))
quantile_RI_AS_6_hclust

#Creating a table with all values from the created quantiles

quantiles_RI_AS_6 <- cbind.data.frame(round(quantile_RI_AS_1_hclust, 2), round(quantile_RI_AS_2_hclust, 2), round(quantile_RI_AS_3_hclust, 2), round(quantile_RI_AS_4_hclust, 2), round(quantile_RI_AS_5_hclust, 2), round(quantile_RI_AS_6_hclust, 2))

alt_RI_1 <- sprintf("Alternitavely Spliced 1 - RI - %d/%d events", nrow(psi_RI_AS_1_hclust), nrow(events_RI_hclust_AltSpl))
alt_RI_2 <- sprintf("Alternitavely Spliced 2 - RI - %d/%d events", nrow(psi_RI_AS_2_hclust), nrow(events_RI_hclust_AltSpl))
alt_RI_3 <- sprintf("Alternitavely Spliced 3 - RI - %d/%d events", nrow(psi_RI_AS_3_hclust), nrow(events_RI_hclust_AltSpl))
alt_RI_4 <- sprintf("Alternitavely Spliced 4 - RI - %d/%d events", nrow(psi_RI_AS_4_hclust), nrow(events_RI_hclust_AltSpl))
alt_RI_5 <- sprintf("Alternitavely Spliced 5 - RI - %d/%d events", nrow(psi_RI_AS_5_hclust), nrow(events_RI_hclust_AltSpl))
alt_RI_6 <- sprintf("Alternitavely Spliced 6 - RI - %d/%d events", nrow(psi_RI_AS_6_hclust), nrow(events_RI_hclust_AltSpl))

col_names_RI_AS_6 <- c(alt_RI_1, alt_RI_2, alt_RI_3, alt_RI_4, alt_RI_5, alt_RI_6)


colnames(quantiles_RI_AS_6) <- col_names_RI_AS_6
quantiles_RI_AS_6
table_quantiles_RI_AS_6_hclust <- kable(quantiles_RI_AS_6, format = "html", table.attr = "border=\"1\"", 
                            caption = "Quantiles Alternatively Spliced - RI - hierarchical clustering", align="ccc") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
table_quantiles_RI_AS_6_hclust
save_kable(x=table_quantiles_RI_AS_6_hclust, file= "/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/quantiles_RI_AS_6_hclust.html")

```


