---
title: "consensus_clustering_SE"
output: html_document
date: "2023-05-15"
author: "Aïsha Schuhegger"
note: "You need to execute the scripts hclust-SE.Rmd and hclust-RI.Rmd before running this."
---

```{r}
library(ConsensusClusterPlus)
```
Parameters for ConsensusClusterPlus:
- d: distance matrix 
- pItem: 80% item resampling
- pFeature: 80% gene resampling 
- maxK: a maximum evaluated k of 6 so that cluster counts of 2,3,4,5,6 are evaluated (in practice a higher cluster count such as 20 is recommended)
- reps: 50 resamplings (! in practice a much hihger reps is recommended such as 1000)
- clusterAlg: agglomerative hierarchical clustering algorithm
- distance: 1-Perason correlation distance
- title: title for the output 

# Skipped Exon
```{r}
dist_dt_SE_rcName <- dist_dt_SE
rownames(dist_dt_SE_rcName) <- rownames(events_SE)
colnames(dist_dt_SE_rcName) <- rownames(events_SE)
dist_SE_cons <- as.dist(as.matrix(dist_dt_SE_rcName))
cons_mat_SE <- ConsensusClusterPlus(dist_SE_cons, maxK=6,reps=50,pItem=0.8,pFeature=1, title='Consensus matrix SE', clusterAlg='hc', distance='pearson', seed=1262118388.71279,plot='png')
```

```{r}
#Top five rows and columns of results for k=6
cons_mat_SE[[3]][["consensusMatrix"]][1:5,1:5]

#ConsensusTree - hclust object
cons_mat_SE[[3]][["consensusTree"]]

#ConsenusClass - the sample classifications
cons_mat_SE[[3]][["consensusClass"]][1:5]

```
```{r}
#ml - consensus matrix result
#clrs - colors for cluster


```
After executing ConsensusClusterPlus, one can optionally calculate cluster-
consensus and item-consensus results by:
```{r}
icl_SE <- calcICL(cons_mat_SE,title='Consensus Clustering SE',plot="png")

icl_SE
icl_SE[["clusterConsensus"]]
```

