---
title: "Plot Output STITCHIT"
output: html_document
author: "Aïsha Schuhegger"
date: "2023-07-15"
---

```{r}
library(tidyverse)
library(dplyr)
```


#Data preparation
#k Spearman/ k Pearson -> p-values are caluclated with Spearman/ Pearson and k p-values had a value < alpha -> k is the amount of significant segments for a specific event (row) 
#SegSpearman/ SegPearson -> Stitchit produces 2 output files for one event, one output file that contains values calculated with Spearman and the other one with Pearson
```{r}
# Read the CSV files into separate data frames
histo_H3K27ac_SegSpearman <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K27ac_SegSpearman.csv")
histo_H3K27me3_SegSpearman <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K27me3_SegSpearman.csv")
histo_H3K36me3_SegSpearman <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K36me3_SegSpearman.csv")
histo_H3K4me1_SegSpearman <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K4me1_SegSpearman.csv")
histo_H3K4me3_SegSpearman <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K4me3_SegSpearman.csv")
histo_H3K9me3_SegSpearman <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K9me3_SegSpearman.csv")

histo_H3K27ac_SegSpearman$class <- rep('H3K27ac', nrow(histo_H3K27ac_SegSpearman))
histo_H3K27me3_SegSpearman$class <- rep('H3K27me3', nrow(histo_H3K27me3_SegSpearman))
histo_H3K36me3_SegSpearman$class <- rep('H3K36me3', nrow(histo_H3K36me3_SegSpearman))
histo_H3K4me1_SegSpearman$class <- rep('H3K4me1', nrow(histo_H3K4me1_SegSpearman))
histo_H3K4me3_SegSpearman$class <- rep('H3K4me3', nrow(histo_H3K4me3_SegSpearman))
histo_H3K9me3_SegSpearman$class <- rep('H3K9me3', nrow(histo_H3K9me3_SegSpearman))
```


```{r}
histo_H3K27ac_SegPearson <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K27ac_SegPearson.csv")
histo_H3K27me3_SegPearson <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K27me3_SegPearson.csv")
histo_H3K36me3_SegPearson <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K36me3_SegPearson.csv")
histo_H3K4me1_SegPearson <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K4me1_SegPearson.csv")
histo_H3K4me3_SegPearson <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K4me3_SegPearson.csv")
histo_H3K9me3_SegPearson <- read_csv("/nfs/proj/splicingREMs/output_Correlation/histo_H3K9me3_SegPearson.csv")

histo_H3K27ac_SegPearson$class <- rep('H3K27ac', nrow(histo_H3K27ac_SegPearson))
histo_H3K27me3_SegPearson$class <- rep('H3K27me3', nrow(histo_H3K27me3_SegPearson))
histo_H3K36me3_SegPearson$class <- rep('H3K36me3', nrow(histo_H3K36me3_SegPearson))
histo_H3K4me1_SegPearson$class <- rep('H3K4me1', nrow(histo_H3K4me1_SegPearson))
histo_H3K4me3_SegPearson$class <- rep('H3K4me3', nrow(histo_H3K4me3_SegPearson))
histo_H3K9me3_SegPearson$class <- rep('H3K9me3', nrow(histo_H3K9me3_SegPearson))
```


#Check if the column 'Event' is the same for all 6 dataframes (Seg Spearman) 
#After execution the first part 'Data preparation' should run again
```{r}
histo_H3K27ac_SegSpearman$Event <- sapply(strsplit(as.character(histo_H3K27ac_SegSpearman$Event), "_H3"), `[`, 1)
histo_H3K27me3_SegSpearman$Event <- sapply(strsplit(as.character(histo_H3K27me3_SegSpearman$Event), "_H3"), `[`, 1)
histo_H3K36me3_SegSpearman$Event <- sapply(strsplit(as.character(histo_H3K36me3_SegSpearman$Event), "_H3"), `[`, 1)
histo_H3K4me1_SegSpearman$Event <- sapply(strsplit(as.character(histo_H3K4me1_SegSpearman$Event), "_H3"), `[`, 1)
histo_H3K4me3_SegSpearman$Event <- sapply(strsplit(as.character(histo_H3K4me3_SegSpearman$Event), "_H3"), `[`, 1)
histo_H3K9me3_SegSpearman$Event <- sapply(strsplit(as.character(histo_H3K9me3_SegSpearman$Event), "_H3"), `[`, 1)


all_same_event <- all(
  identical(histo_H3K27ac_SegSpearman$Event, histo_H3K27me3_SegSpearman$Event),
  identical(histo_H3K27ac_SegSpearman$Event, histo_H3K36me3_SegSpearman$Event),
  identical(histo_H3K27ac_SegSpearman$Event, histo_H3K4me1_SegSpearman$Event),
  identical(histo_H3K27ac_SegSpearman$Event, histo_H3K4me3_SegSpearman$Event),
  identical(histo_H3K27ac_SegSpearman$Event, histo_H3K9me3_SegSpearman$Event)
)

if(all_same_event){
  print("Yes, the column 'Event' is identical between all dataframes (regarding SegSpearman).")
} else {
  print("No, the column 'Event' is not identical between all dataframes (regarding SegSpearman).")
}
```

#Check if the column 'Event' is the same for all 6 dataframes (Seg Pearson) 
#After execution the first part 'Data preparation' should run again
```{r}
histo_H3K27ac_SegPearson$Event <- sapply(strsplit(as.character(histo_H3K27ac_SegPearson$Event), "_H3"), `[`, 1)
histo_H3K27me3_SegPearson$Event <- sapply(strsplit(as.character(histo_H3K27me3_SegPearson$Event), "_H3"), `[`, 1)
histo_H3K36me3_SegPearson$Event <- sapply(strsplit(as.character(histo_H3K36me3_SegPearson$Event), "_H3"), `[`, 1)
histo_H3K4me1_SegPearson$Event <- sapply(strsplit(as.character(histo_H3K4me1_SegPearson$Event), "_H3"), `[`, 1)
histo_H3K4me3_SegPearson$Event <- sapply(strsplit(as.character(histo_H3K4me3_SegPearson$Event), "_H3"), `[`, 1)
histo_H3K9me3_SegPearson$Event <- sapply(strsplit(as.character(histo_H3K9me3_SegPearson$Event), "_H3"), `[`, 1)


all_same_event <- all(
  identical(histo_H3K27ac_SegPearson$Event, histo_H3K27me3_SegPearson$Event),
  identical(histo_H3K27ac_SegPearson$Event, histo_H3K36me3_SegPearson$Event),
  identical(histo_H3K27ac_SegPearson$Event, histo_H3K4me1_SegPearson$Event),
  identical(histo_H3K27ac_SegPearson$Event, histo_H3K4me3_SegPearson$Event),
  identical(histo_H3K27ac_SegPearson$Event, histo_H3K9me3_SegPearson$Event)
)

if(all_same_event){
  print("Yes, the column 'Event' is identical between all dataframes (regarding SegPearson).")
} else {
  print("No, the column 'Event' is not identical between all dataframes (regarding SegPearson).")
}
```

#Create combined dataframe 
```{r}
combined_histo_HM_SegSpearman <- bind_rows(histo_H3K27ac_SegSpearman, histo_H3K27me3_SegSpearman, histo_H3K36me3_SegSpearman, histo_H3K4me1_SegSpearman, histo_H3K4me3_SegSpearman, histo_H3K9me3_SegSpearman)
combined_histo_HM_SegPearson <- bind_rows(histo_H3K27ac_SegPearson, histo_H3K27me3_SegPearson, histo_H3K36me3_SegPearson, histo_H3K4me1_SegPearson, histo_H3K4me3_SegPearson, histo_H3K9me3_SegPearson)
```

#Plots for Segmentation_..._Spearman.txt input files 
```{r}
#TODO: Explain kSpearman and SegSpearman in the comments and/or plot!
#TODO: Add title to the plots 

pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/6histograms_kSpearman_SegSpearman.pdf')
plot_kSpearman_SegSpearman <- ggplot(combined_histo_HM_SegSpearman, aes(x=combined_histo_HM_SegSpearman$`k Spearman`)) + 
                              geom_histogram(colour="black", fill="lightblue") +
                              xlab("Amount of Significant Segments within an Event") +
                              ylab("Count (Amount of Events)") + 
                              labs(title="Amount of Significant Segments (p-value calculated with Spearman) & Segmentation_..._Spearman.txt files as input files") + 
                              facet_wrap(vars(class))
dev.off()

pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/6histograms_meanLengthSpearman_SegSpearman.pdf')
plot_meanLengthSpearman_SegSpearman <- ggplot(combined_histo_HM_SegSpearman, aes(x=combined_histo_HM_SegSpearman$`mean significant segments length Spearman`)) + 
                              geom_histogram(colour="black", fill="lightblue") +
                              xlab("Mean length of the significant segments within an Event") +
                              ylab("Count (Amount of Events)") + 
                              labs(title="Mean Length of the Significant Segments (p-value calculated with Spearman) & Segmentation_..._Spearman.txt files a sinput files") + 
                              facet_wrap(vars(class))
dev.off()

pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/6histograms_kPearson_SegSpearman.pdf')
plot_kPearson_SegSpearman <- ggplot(combined_histo_HM_SegSpearman, aes(x=combined_histo_HM_SegSpearman$`k Pearson`)) + 
                              geom_histogram(colour="black", fill="lightblue") +
                              xlab("Amount of Significant Segments within an Event") +
                              ylab("Count (Amount of Events)") + 
                              labs(title="Amount of Significant Segments (p-value calculated with Pearson) & Segmentation_..._Spearman.txt files as input files") + 
                              facet_wrap(vars(class))
dev.off()

pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/6histograms_meanLengthPearson_SegSpearman.pdf')
plot_meanLengthPearson_SegSpearman <- ggplot(combined_histo_HM_SegSpearman, aes(x=combined_histo_HM_SegSpearman$`mean significant segments length Pearson`)) + 
                              geom_histogram(colour="black", fill="lightblue") +
                              xlab("Mean length of the significant segments within an Event") +
                              ylab("Count (Amount of Events)") + 
                              labs(title="Mean Length of the Significant Segments (p-value calculated with Pearson) & Segmentation_..._Spearman.txt files a sinput files") + 
                              facet_wrap(vars(class))
dev.off()
```


#Plots for Segmentation_..._Pearson.txt input files
```{r}
pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/6histograms_kSpearman_SegPearson.pdf')
plot_kSpearman_SegPearson <- ggplot(combined_histo_HM_SegPearson, aes(x=combined_histo_HM_SegPearson$`k Spearman`)) + 
                              geom_histogram(colour="black", fill="lightpink") +
                              xlab("Amount of Significant Segments within an Event") +
                              labs(title="Amount of Significant Segments (p-value calculated with spearman) & Segmentation_..._Pearson.txt files as input files") +
                              ylab("Count (Amount of Events)") + 
                              facet_wrap(vars(class))
dev.off()

pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/6histograms_meanLengthSpearman_SegPearson.pdf')
plot_meanLengthSpearman_SegPearson <- ggplot(combined_histo_HM_SegPearson, aes(x=combined_histo_HM_SegPearson$`mean significant segments length Spearman`)) + 
                              geom_histogram(colour="black", fill="lightpink") +
                              xlab("Mean length of the significant segments within an Event") +
                              ylab("Count (Amount of Events)") + 
                              labs(title="Mean Length of the Significant Segments (p-value calculated with Spearman) & Segmentation_..._Pearson.txt files a sinput files") +
                              facet_wrap(vars(class))
dev.off()
                              
pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/6histograms_kPearson_SegPearson.pdf')
plot_kPearson_SegPearson <- ggplot(combined_histo_HM_SegPearson, aes(x=combined_histo_HM_SegPearson$`k Pearson`)) + 
                              geom_histogram(colour="black", fill="lightpink") +
                              xlab("Amount of Significant Segments within an Event") +
                              labs(title="Amount of Significant Segments (p-value calculated with Pearson) & Segmentation_..._Pearson.txt files as input files") +
                              ylab("Count (Amount of Events)") + 
                              facet_wrap(vars(class))
dev.off()

pdf('/nfs/home/students/a.schuhe/scripts/splicingREMs/plots/6histograms_meanLengthPearson_SegPearson.pdf')
plot_meanLengthPearson_SegPearson <- ggplot(combined_histo_HM_SegPearson, aes(x=combined_histo_HM_SegPearson$`mean significant segments length Pearson`)) + 
                              geom_histogram(colour="black", fill="lightpink") +
                              xlab("Mean length of the significant segments within an Event") +
                              ylab("Count (Amount of Events)") + 
                              labs(title="Mean Length of the Significant Segments (p-value calculated with Pearson) & Segmentation_..._Pearson.txt files a sinput files") + 
                              facet_wrap(vars(class))
dev.off()
```


